---
title: "Bayesian Re-analysis (Staircases)"
date: today
author: "Imane Lattab"
institute: "The University of Sydney"
repository: https://github.com/imanel04/PsychHonours.git
format: 
  html:
    embed-resources: true
    code-fold: true
    code-tools: true
    theme: sandstone
    fig_caption: yes
    table-of-contents: true
    toc: true
    toc_depth: 4
    toc_float: true
execute:
  echo: true
  tidy: true
number-sections: false 
---

```{r}
#| label: setup
#| include: false

library("dplyr")
library("ggplot2")
library("patchwork")

# MIXED MODELS
library("lme4") # mle
library("lmerTest")

# BAYESIAN
library("rstan")
library("tidybayes")

library("logspline") # density estimation
library("bridgesampling") # bayes factors
```

# July 2026 Workflow

## Multiple Staircases

Cottier, T. V., Turner, W., Holcombe, A. O., & Hogendoorn, H. (2023). Exploring the extent to which shared mechanisms contribute to motion-position illusions. Journal of Vision, 23(10), 8. https://doi.org/10.1167/jov.23.10.8

### Data

```{r}
FLE_data <- read.csv("data/Cottier2023/FLE_Raw/all_participants_staircase_FLEs.csv") # check zip! in Cottier2023/FLE_Raw

# ID, session, and staircase as categorical
FLE_data <- FLE_data |>
  dplyr::mutate(
    ID = factor(ID),
    session = factor(session),
    staircase = factor(staircase)
  )
```

Estimate overall mean FLE and participant specific deviations

```{r}
fle_lmer <- lmer(FLE_dva ~ 1 + (1 | ID), data = FLE_data)

summary(fle_lmer)
```

```{r}
mle_p_negative <- pnorm(
  q = 0
  , mean = fixef(fle_lmer)[1]
  , sd = attr(VarCorr(fle_lmer)$ID, "stddev")
)

mle_p_negative
```

Notice that this is larger than the data from the original "session 1 and 2" dataset because I realised later on that there are actually more participants in this dataset that was shared with me.

---

## Bayesian Hierarchical Model

I am just applying the framework from `flash-lag-effect-extended.html`

### Prior Distributions

```{r}
p_negative_prior <- rnorm(1e5, mean = -2/3, sd = 2/3) |>
  pnorm(0) |>
  (\(x) x / 2)()

p_negative_prior |>
  logspline::logspline(lbound = 0, ubound = 0.5) |>
  (
    \(fit) {
      curve(
        logspline::dlogspline(x, fit)
        , from = 0, to = 0.5
        , lwd = 2
        , xlab = expression(p[theta[i] < 0])
        , ylab = "Density"
        , las = 1
        , n = 1001
      )
    }
  )()
```

```{r}
curve(
  dlnorm(x, meanlog = 0.5, sdlog = 1)
  , from = 0, to = 5
  , lwd = 2
  , xlab = expression(mu[theta])
  , ylab = "Density"
  , las = 1
  , n = 1001
)
```

```{r}
mu_prior <- rlnorm(1e5, meanlog = 0.5, sdlog = 1)
sigma_prior <- -mu_prior / qnorm(p_negative_prior)

sigma_prior |>
  logspline::logspline(lbound = 0) |>
  (
    \(fit) {
      curve(
        logspline::dlogspline(x, fit)
        , from = 0, to = 5
        , lwd = 2
        , xlab = expression(sigma[theta])
        , ylab = "Density"
        , las = 1
        , n = 1001
      )
    }
  )()
```

```{r}
unconstrained_joint_prior <- data.frame(mu = mu_prior, sigma = sigma_prior, p_negative = p_negative_prior) |>
  ggplot() +
    aes(x = mu, y = sigma, color = p_negative) +
    geom_point(alpha = 0.1) +
    stat_density_2d(color = "white", bins = 10) +
    geom_abline(intercept = 0, slope = 1, color = "white", linetype = "22") +
    # annotate("point", x = fixef(fle_lmer)[1], y = attr(VarCorr(fle_lmer)$ID, "stddev"), shape = 21, color = "black", fill = "white", size = 5) +
    labs(
      x = expression(mu[theta])
      , y = expression(sigma[theta])
      , color = expression(p(theta[i] < 0))
    ) +
    scale_color_viridis_c(option = "C", limits = c(0, 0.5)) +
    lims(x = c(0, 5), y = c(0, 5)) +
    papaja::theme_apa(base_size = 16)

unconstrained_joint_prior
```


### Model Estimation

```{r}
#| label: fit-bayesian-model

# Read the Stan model
pse_model <- readLines(
  "models/pse-unconstrained.stan"
) |>
  paste(collapse = "\n") # model files are from in this github repo https://github.com/alexholcombe/qualitativeIndividDiffsMotnPostnIllusions.git

# Prepare staircase-level data for Stan
FLE_data_model <- FLE_data |>
  filter(
    !is.na(ID),
    !is.na(FLE_dva)
  ) |>
  mutate(
    ID = as.factor(ID),
    subject_id = as.integer(ID)
  ) |>
  arrange(subject_id, session, staircase)

# Check number of observations per participant
FLE_data_model |>
  count(ID, name = "n_observations") |>
  count(n_observations)

fle_stan_data <- list(
  # Total number of staircase-level FLE estimates
  N = as.integer(nrow(FLE_data_model)),

  # Number of unique participants
  n_subjects = as.integer(n_distinct(FLE_data_model$ID)),

  # Participant index for every observation
  id = as.integer(FLE_data_model$subject_id),

  # Staircase-level FLE estimates in DVA
  y = as.numeric(FLE_data_model$FLE_dva),

  # Prior settings
  pi_mean = -2 / 3,
  pi_sd = 2 / 3,
  mu_theta_mean = 0.5,
  mu_theta_sd = 1
)

stopifnot(
  fle_stan_data$N == length(fle_stan_data$y),
  fle_stan_data$N == length(fle_stan_data$id),
  fle_stan_data$n_subjects == length(unique(fle_stan_data$id)),
  all(is.finite(fle_stan_data$y)),
  min(fle_stan_data$id) == 1L,
  max(fle_stan_data$id) == fle_stan_data$n_subjects
)

str(fle_stan_data)

fle_bayes <- rstan::stan(
  model_code = pse_model,
  data = fle_stan_data,
  iter = 1e5 / 4 + 1000,
  warmup = 1000,
  chains = 4,
  cores = 4,
  seed = 123
)

# Model diagnostics
str_post_unconstrained <- data.frame(
  summary(fle_bayes)$summary
)

rstan::check_hmc_diagnostics(fle_bayes)
```
```{r}
#| label: tbl-compare-estimates
#| tbl-cap: "Comparison of Bayesian posterior means and MLE estimates."

data.frame(
  parameter = c("mu_theta", "sigma_theta", "sigma_epsilon", "p_negative_theta")
  , bayes = fle_bayes |>
    summary(pars = c("mu_theta", "sigma_theta", "sigma_epsilon", "p_negative_theta")) |>
    (\(x) x$summary)() |>
    as.data.frame() |>
    dplyr::pull("mean")
  , mle = c(
    fixef(fle_lmer)[1],
    attr(VarCorr(fle_lmer)$ID, "stddev"),
    sigma(fle_lmer),
    mle_p_negative
  )
) |>
  knitr::kable()
```

```{r}
#| label: bayesian-model-summary

fle_bayes |>
  summary(pars = c("mu_theta", "sigma_theta", "sigma_epsilon", "p_negative_theta")) |>
  (\(x) x$summary)() |>
  tidyr::as_tibble(rownames = "parameter") |>
  dplyr::mutate(
    parameter = gsub("mu", "$\\\\mu", parameter) |>
      gsub("sigma", "$\\\\sigma", x = _) |>
      gsub("_(.+)", "_{\\\\\\1}$", x = _) |>
      gsub("^p.+", "$p(\\\\theta_i<0)$", x = _)
  ) |>
  knitr::kable(digits = 3)
```

```{r}
#| label: fig-posterior-distributions
#| fig-cap: "Posterior distributions for the group mean of individual differences $\\mu_{\\theta}$, the standard deviation of individual differences $\\sigma_{\\theta}$, and the proportion of participants with a negative flash lag effect $p(\\theta_i<0)$."

(tidybayes::gather_draws(
  fle_bayes
  , mu_theta, sigma_theta, p_negative_theta
) |>
  dplyr::mutate(
    .variable = factor(
      .variable
      , levels = c("mu_theta", "sigma_theta", "p_negative_theta")
      , labels = c("mu[theta]", "sigma[theta]", "p(theta[i] < 0)")
    )
  ) |>
  ggplot() +
  aes(x = .value, fill = after_stat(level)) +
  stat_halfeye(fill_type = "segments", normalize = "panels", .width = c(.5, .8, .95, .99, 1), slab_color = "black", alpha = 0.7, fatten_point = 2, interval_size_range = c(1, 2)) +
  annotate("point", x = c(0, 0.5), y = 0, color = "white") + 
  scale_fill_viridis_d(option = "B", begin = 0.9, end = 0.1) +
  guides(
    fill = guide_legend(
      title = "Credible level"
      , override.aes = list(color = NA)
      , reverse = TRUE
      , direction = "horizontal"
      , title.position = "top"
      , title.hjust = 0.5
    )
  ) +
  # coord_cartesian(ylim = c(-0.05, NA)) +
  facet_wrap(
    ~ .variable
    , scales = "free"
    , labeller = label_parsed
    , ncol = 2
  ) +
  labs(x = "Parameter value") +
  papaja::theme_apa(base_size = 16) +
  theme(
    axis.line.y = element_blank()
    , axis.ticks.y = element_blank()
    , axis.text.y = element_blank()
    , axis.title.y = element_blank()
    # , legend.title.position = "top"
  )) |>
  lemon::reposition_legend("center", panel = 'panel-2-2')
```
Notice that the estimate of negative FLE people is much higher than in the original 2 session dataset; this is likely because this dataset I'm using actually has a lot more participants, which I suspect some of them are potentially low-quality particioants or from other experiments. I will look into cleaning these people out with Tim, but I will continue.

---

### Estimates of individual effects

```{r}
#| label: fig-individual-effects

theta_i <- tidybayes::gather_draws(
  fle_bayes
  , theta[i]
)

# i as factor sorted by median of posterior distribution
theta_i_ordered <- tidybayes::median_qi(theta_i) |>
  mutate(
    i = factor(
      i
      , labels = unique(i)[order(.value)]
      , levels = unique(i)[order(.value)]
    )
  )

theta_caterpillar <- theta_i |>
  dplyr::mutate(
    # i as factor sorted by median of posterior distribution
    ID = factor(
      i
      , labels = unique(theta_i_ordered$i)[order(theta_i_ordered$.value)]
      , levels = unique(theta_i_ordered$i)[order(theta_i_ordered$.value)]
    )
  ) |>
  dplyr::rename(theta = .value) |>
  ggplot() +
    aes(x = ID, y = theta) +
    geom_hline(yintercept = 0, color = "black", linetype = "22") +
    stat_pointinterval(aes(fill = after_stat(y) < 0, interval_color = after_stat(y) < 0), point_color = "white", .width = c(.95), fatten_point = 2, linewidth = 1, shape = 21) +
    scale_fill_viridis_d(option = "B", begin = 0.9, end = 0.1, aesthetics = c("fill", "interval_color")) +
    guides(interval_color = "none", fill = "none", linewidth = "none") +
    labs(y = expression(theta), x = "Participant ID") +
    coord_cartesian(ylim = c(-5, 7.5)) +
    papaja::theme_apa(base_size = 16) +
    theme(
      # axis.line.x = element_blank()
      # , axis.ticks.x = element_blank()
      , axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5, hjust = 1)
    )
```

```{r}
#| label: fig-posterior-population-distribution

posterior_population_distribution <- tidybayes::gather_draws(
  fle_bayes
  , mu_theta, sigma_theta
) |>
  tidyr::nest(data = c(.chain, .iteration, .draw)) |>
  tidyr::pivot_wider(names_from = .variable, values_from = .value) |>
  dplyr::mutate(
    density = purrr::map2(
      mu_theta
      , sigma_theta
      , \(mu, sigma) {
        dnorm(
          x = seq(-5, 7.5, length.out = 101)
          , mean = mu
          , sd = sigma
        )
      }
    )
    , theta = list(seq(-5, 7.5, length.out = 101))
  )

theta_dist <- posterior_population_distribution |>
  tidyr::unnest(cols = c(density, theta)) |>
  ggplot() +
    aes(x = theta, y = density) +
    geom_vline(xintercept = 0, color = "black", linetype = "22") +
    geom_hline(yintercept = 0, color = "black") +
    geom_lineribbon(
      aes(ymin = after_stat(ymin), ymax = after_stat(ymax))
      , stat = "summary"
      , fun.data = "mean_qi"
      , color = "black"
      , fill = "black"
      , alpha = 0.2
    ) +
    # geom_rug(
    #   data = tidybayes::median_qi(theta_i)
    #   , aes(x = .value, y = 0)
    #   , sides = "b"
    #   , color = "black"
    #   , alpha = 0.5
    # ) +
    # geom_point(
    #   data = tidybayes::median_qi(theta_i)
    #   , aes(x = .value, y = 0, fill = .value < 0)
    #   , color = "white"
    #   , shape = 21
    #   , size = 2
    # ) +
    scale_fill_viridis_d(option = "B", begin = 0.9, end = 0.1) +
    guides(fill = "none") +
    labs(x = expression(theta), y = "Density") +
    papaja::theme_apa(base_size = 16)
```

```{r}
#| label: fig-posterior-theta-distribution
#| fig-cap: "Posterior distribution of individual effects $\\theta_i$ for each participant (left) and the posterior distribution of the population distribution of individual effects $\\theta_i$ (right). The horizontal dashed line marks the point of no effect, i.e., $\\theta_i = 0$."

theta_caterpillar + (theta_dist + coord_flip() + theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())) +
  plot_layout(ncol = 2, widths = c(1, 1/3), guides = "collect", axes = "collect_y")
```


## Bayes factor approximations

Finally, we compute the Bayes factor in favor of the assumption of an unconstrained population distribution ($p(\theta_i<0) > 0$) over the strictly positive population distribution (stochastic dominance).
One approach is to approximate this Bayes factor using the Savage-Dickey density ratio for $p(\theta_i<0) = 0$.

```{r}
#| label: fig-bayes-factor
#| fig-cap: "Posterior and prior density for the proportion of participants with a negative flash lag effect, $p(\\theta_i<0)$. The Bayes factor in favor of the alternative hypothesis ($p(\\theta_i<0) > 0$) over the null hypothesis ($p(\\theta_i<0) = 0$) is given by the ratio of the prior density to the posterior density at $p(\\theta_i<0) = 0$."

# Extract posterior samples for p_negative_theta
p_negative_draws <- tidybayes::gather_draws(fle_bayes, p_negative_theta)

# logspline approximation of posterior density
p_negative_draws_ls <- logspline::logspline(
  p_negative_draws$.value
  , lbound = 0, ubound = 0.5
)

# logspline approximation of prior density
prior_ls <- rnorm(1e5, mean = fle_stan_data$pi_mean, sd = fle_stan_data$pi_sd) |>
  pnorm() |>
  (\(x) x / 2)() |>
  logspline::logspline(lbound = 0, ubound = 0.5)

prior_density <- logspline::dlogspline(0, prior_ls)
posterior_density <- logspline::dlogspline(0, p_negative_draws_ls)


curve(
  logspline::dlogspline(x, p_negative_draws_ls)
  , from = 0, to = 0.5
  , col = "black", lwd = 3
  , main = "Posterior Density of P(p_negative_theta)"
  , xlab = "p_negative_theta"
  , n = 1001
)

curve(
  logspline::dlogspline(x, prior_ls)
  , from = 0, to = 0.5
  , col = "grey", lwd = 3
  , add = TRUE
  , n = 1001
)

points(
  x = c(0, 0)
  , y = c(prior_density, posterior_density)
  , col = c("white", "white")
  , bg = c("grey", "black")
  , pch = 21
  , cex = 2
)
```

```{r}
#| label: bayes-factor-calculation

bayes_factor_10 <- prior_density / posterior_density
bayes_factor_10
```

We still find overwhelming evidence for the unconstrained population distribution. In other words, the assumption that noone exhibits a flash lead effect, seems highly implausible.

## Stochastic Dominance Analysis

**Positive effect model (dominance model)**: 
Order constraint, constrains individual effects to be positive, but effects differ/vary interindividually. Truncated normal distribution as suggested by literature.

Defining $\sigma_{\theta}$ the same way as in the unconstrained model would not make any sense conceptually, since in this model $p(\theta_i<0)$ is assumed to be zero. Therefore chose log-normal as informative prior (as prior of unconstrained model was similar to prior for $\mu_\theta$).

**Common-effect model**: 
No individual variablity in effects, all inidividual share a common, constant effect different from zero (positive).
Unlikely to be an accurate representation of effects in psychology, but used as (1) benchmark to check for individual differences (model could be preferred should individual difference be very small) and (2) design check (design might not be able to capture individual differences).

Priors are only needed for $\theta$ (as there is uncertainty regarding the size of the common effect) and $\sigma_\epsilon^2$. Logarithmic normal distribution as prior for $\theta$, constraining it to positive real numbers (paralleling prior of $\mu_{\theta}$ in unconstrained model). 

**Average null model**:
Average effect across individuals is constrained to be zero, but effects are assumed to vary across individuals. Conceptualized as a more forgiving alternative to the null model (see below), less reactive to contaminated data. $\theta_i$ are modeled as normally distributed around a mean of zero. Only slight deviations from zero align with the idea of this model as an alternative to the null model, therefore steep log-normal distribution with relatively small mean to be chosen as prior for $\sigma_{\theta}$ (see average null vs unconstrained).

**Null model**: 
Specifies a true effect of zero for all individuals, most constrained model. Used as benchmark for no effect and no interindividual variability. 
There is no uncertainty regarding the size of the effect, as it is assumed to be zero. So no prior for theta. Only prior for sigma (measurement error around true $\theta = 0$) needed. 

---

## (1) Dominance (truncated normal) vs unconstrained

Priors chosen to be similar to priors of unconstrained model, therefore approximately same percentage of theoretically negative $\theta_i$. But this time, negative $\theta_i$ are excluded / impossible by stan model definition (truncated normal distribution).

```{r}
#| label: prior-plot-mu-dominance
#| fig-cap: "Prior distribution of $\\mu_{\\theta}$ in dominance model"

mu_prior_truncn <- rlnorm(1e5, meanlog = 0.5, sdlog = 1)

curve(
  dlnorm(x, meanlog = 0.5, sdlog = 1)
  , from = 0, to = 5
  , lwd = 2
  , xlab = expression(mu[theta])
  , ylab = "Density"
  , las = 1
  , n = 1001
)
```

```{r}
#| label: prior-plot-sigma-dominance
#| fig-cap: "Prior distribution of $\\sigma_{\\theta}$ in dominance model"

sigma_prior_truncn <- rlnorm(1e5, meanlog = 0.5, sdlog = 1)

curve(
  dlnorm(x, meanlog = 0.5, sdlog = 1)
  , from = 0, to = 5
  , lwd = 2
  , xlab = expression(sigma[theta])
  , ylab = "Density"
  , las = 1
  , n = 1001
)
```

```{r}
#| label: fit-dominance-model

pse_dominance_model <- readLines("models/pse-dominance.stan") |>
  paste(collapse = "\n")

fle_stan_dominance_data <- list(
  N = nrow(FLE_data) |> as.integer()
  , n_subjects = length(unique(FLE_data$ID)) |> as.integer()
  , id = as.integer(as.factor(FLE_data$ID))
  , y = FLE_data$FLE_dva

  # Prior settings
  , mu_theta_mean = 0.5
  , mu_theta_sd = 1
  , sigma_theta_mean = 0.5
  , sigma_theta_sd = 1
)

fle_dominance_bayes <- rstan::stan(
  model_code = pse_dominance_model
  , data = fle_stan_dominance_data
  , iter = 1e5/4 + 1000
  , warmup = 1000
  , chains = 4
  , cores = 4
  , seed = 123
)

# model diagnostics
str_post_dominance<- data.frame(summary(fle_dominance_bayes)$summary)  
check_hmc_diagnostics(fle_dominance_bayes)

```


### Estimates of individual effects

```{r}
#| label: fig-indiv-effects-truncated

theta_i_truncn <- tidybayes::gather_draws(
  fle_dominance_bayes
  , theta[i]
)

theta_i_trunc_ordered <- tidybayes::median_qi(theta_i_truncn) |>
  mutate(i = factor(i
                    ,levels = unique(i)[order(.value)]        #levels are ordered after median, 64 = min value = first level
                    ,labels = unique(i)[order(.value)]        #I create this solely for creating levels in draws, as geom-f calls on draws
                    ) 
         )

# Plot median estimates
truncated_caterpillar <- theta_i_truncn |>
  mutate(ID = factor(i
                     ,levels = unique(theta_i_trunc_ordered$i)[order(theta_i_trunc_ordered$.value)]  #new var ID with ordered med levels
                     ,labels = unique(theta_i_trunc_ordered$i)[order(theta_i_trunc_ordered$.value)]
                     )
  ) |>
  rename(theta = .value) |>
  ggplot() +
  aes(x = ID, y = theta) +
  geom_hline(yintercept = 0, color = "black", linetype = "22") + #string-linetype: line for 2 units, followed by 2 off
  stat_pointinterval(           #draws point est + CI for draw-input, point_interval function is median_qi by default
    fill = viridisLite::inferno(1, begin = 0.9), interval_color = viridisLite::inferno(1, begin = 0.9), point_color = "white"
    , .width = c(.95), fatten_point = 2, linewidth = 1, shape = 21
  ) +
  labs(y = expression(theta[i]), x = "Participant ID") +
  coord_cartesian(ylim = c(0,6)) +
  papaja::theme_apa(base_size = 16) + 
  theme(
    axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5, hjust = 0)
  )
  
```

```{r}
#| label: fig-post-population-distribution-truncated

post_populDist_truncated <- tidybayes::gather_draws(
  fle_dominance_bayes
  , mu_theta, sigma_theta
) |>
  tidyr::nest(data = c(.chain, .iteration, .draw)) |>
  tidyr::pivot_wider(names_from = .variable, values_from = .value) |>
  mutate(
    density = purrr::map2(                      # iteration of each density function for each pair of mu and sigma
      mu_theta, sigma_theta, 
      \(mu, sigma){
        dnorm(
          x = seq(0, 6, length.out = 101)
          , mean = mu
          , sd = sigma
        )
      }
    )
    ,theta = list(seq(0, 6, length.out = 101))    # corresponsing values on x for plot
  )
  
 theta_trunc_dist <- post_populDist_truncated |>
  tidyr::unnest(cols = c(density, theta)) |>
  ggplot() + 
  aes(x = theta, y = density) +
   geom_vline(xintercept = 0, color = "black", linetype = "22") +
   geom_hline(yintercept = 0, color = "black") + 
   geom_lineribbon(                              # function designed to use with output from point_interval
     aes(ymin = after_stat(ymin), ymax = after_stat(ymax))
     , stat = "summary"
     , fun.data = "median_qi"
     , color = "black"
     , fill = "black"
     , alpha = 0.2
   ) + 
   labs(x = expression(theta[i]), y = "Density") + 
   papaja::theme_apa(base_size = 16)

```

```{r}
#| label: fig-posterior-theta-distribution-truncated
#| fig-cap: "Posterior distribution of individual effects $\\theta_i$ for each participant (left) and the posterior distribution of the population distribution of individual effects $\\theta_i$ (right). The horizontal dashed line marks the point of no effect, i.e., $\\theta_i = 0$."

truncated_caterpillar + (theta_trunc_dist + coord_flip() + theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())) +
  plot_layout(ncol = 2, width = c(1, 1/3), axes = "collect_y")
```


### Bayes factor with unconstrained model

```{r}
#| label: bayes-factor-dominance

unconstrained_ml <- bridge_sampler(fle_bayes, method = "warp3", repetitions = 30, silent = TRUE)
dominance_ml <- bridge_sampler(fle_dominance_bayes, method = "warp3", repetitions = 30, silent = TRUE)

# bayes factor
bayes_factor(unconstrained_ml, dominance_ml)

```
---


## (2) Common vs unconstrained

```{r}
#| label: prior-plot-common
#| fig-cap: "Prior distribution of $\\theta$ in common model"

mu_prior_common <- rlnorm(1e5, meanlog = 0.5, sdlog = 1)

curve(
  dlnorm(x, meanlog = 0.5, sdlog = 1)
  , from = 0, to = 5
  , lwd = 2
  , xlab = expression(theta)
  , ylab = "Density"
  , las = 1
  , n = 1001
)

```

```{r}
#| label: fit-common-model

pse_common_model <- readLines("models/pse-common.stan") |>
  paste(collapse = "\n")

fle_stan_common_data <- list(
  N = nrow(FLE_data) |> as.integer()
  , n_subjects = length(unique(FLE_data$ID)) |> as.integer()
  , id = as.integer(as.factor(FLE_data$ID))
  , y = FLE_data$FLE_dva

  # Prior settings
  , theta_mean = 0.5
  , theta_sd = 1
)

fle_common_bayes <- rstan::stan(
  model_code = pse_common_model
  , data = fle_stan_common_data
  , iter = 1e5/4 + 1000
  , warmup = 1000
  , chains = 4
  , cores = 4
  , seed = 123
)

# model diagnostics
str_post_common <- data.frame(summary(fle_common_bayes)$summary)
check_hmc_diagnostics(fle_common_bayes)
```

```{r}
#| label: fig-posterior-distributions-common
#| fig-cap: "Posterior distributions for the common effect $\\theta$."

tidybayes::gather_draws(
  fle_common_bayes
  , theta
) |>
  ggplot() +
  aes(x = .value, fill = after_stat(level)) +
  stat_halfeye(fill_type = "segments", normalize = "panels", .width = c(.5, .8, .95, .99, 1), slab_color = "black", alpha = 0.7, fatten_point = 2, interval_size_range = c(1, 2)) +
  annotate("point", x = c(0, 0.5), y = 0, color = "white") + 
  scale_fill_viridis_d(option = "B", begin = 0.9, end = 0.1) +
  guides(
    fill = guide_legend(
      title = "Credible level"
      , override.aes = list(color = NA)
      , reverse = TRUE
      , direction = "horizontal"
      , title.position = "top"
      , title.hjust = 0.5
    )
  ) +
  labs(x = expression(theta), y = "Density") +
  papaja::theme_apa(base_size = 16)

```

### Bayes factor with unconstrained model

```{r}
#| label: bayes-factor-common

common_ml <- bridge_sampler(fle_common_bayes, method = "warp3", repetitions = 30, silent = TRUE)

bayes_factor(unconstrained_ml, common_ml)

```

---

## (3) Average null vs unconstrained

```{r}
#| label: prior-plot-average
#| fig-cap: "Prior distribution of $\\sigma_{\\theta}$ in average null model"

sigma_prior_average <- rlnorm(1e5, meanlog = -0.5, sdlog = 0.5)

curve(
  dlnorm(x, meanlog = -0.5, sdlog = 0.5)
  , from = 0, to = 5
  , lwd = 2
  , xlab = expression(sigma[theta])
  , ylab = "Density"
  , las = 1
  , n = 1001
)

# Mean of prior sigma_theta
exp(-0.5 + (0.5)^2/2)

```

```{r}
#| label: fit-average-model

pse_average_model <- readLines("models/pse-average-null.stan") |>
  paste(collapse = "\n")

fle_stan_average_data <- list(
  N = nrow(FLE_data) |> as.integer()
  , n_subjects = length(unique(FLE_data$ID)) |> as.integer()
  , id = as.integer(as.factor(FLE_data$ID))
  , y = FLE_data$FLE_dva

  # Prior settings
  , sigma_theta_mean = -0.5
  , sigma_theta_sd = 0.5
)

fle_average_bayes <- rstan::stan(
  model_code = pse_average_model
  , data = fle_stan_average_data
  , iter = 1e5/4 + 1000
  , warmup = 1000
  , chains = 4
  , cores = 4
  , seed = 123
)

# model diagnostics
str_post_average <- data.frame(summary(fle_average_bayes)$summary)
check_hmc_diagnostics(fle_average_bayes)
```

### Estimates of individual effects

```{r}
#| label: fig-indiv-effects-average

theta_i_average <- tidybayes::gather_draws(
  fle_average_bayes
  , theta[i]
)

theta_i_average_ordered <- tidybayes::median_qi(theta_i_average) |>
  mutate(i = factor(i
                    ,levels = unique(i)[order(.value)]        
                    ,labels = unique(i)[order(.value)]        #I create this solely for creating levels in draws, as geom-f calls on draws
                    ) 
         )

# Plot median estimates
average_caterpillar <- theta_i_average |>
  mutate(ID = factor(i
                     ,levels = unique(theta_i_average_ordered$i)[order(theta_i_average_ordered$.value)]  
                     ,labels = unique(theta_i_average_ordered$i)[order(theta_i_average_ordered$.value)]
                     )
  ) |>
  rename(theta = .value) |>
  ggplot() +
  aes(x = ID, y = theta) +
  geom_hline(yintercept = 0, color = "black", linetype = "22") + #string-linetype: line for 2 units, followed by 2 off
  stat_pointinterval(           #draws point est + CI for draw-input, point_interval function is median_qi by default
    aes(fill = after_stat(y) < 0, interval_color = after_stat(y) < 0), 
    point_color = "white", .width = c(.95), fatten_point = 2, linewidth = 1, shape = 21
  ) +
  scale_fill_viridis_d(option = "B", begin = 0.9, end = 0.1, aesthetics = c("fill", "interval_color")) +
  guides(interval_color = "none", fill = "none", linetype = "none") +
  labs(y = expression(theta[i]), x = "Participant ID") +
  coord_cartesian(ylim = c(-4,7)) +
  papaja::theme_apa(base_size = 16) + 
  theme(
    axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5, hjust = 0)
  )
  
```

```{r}
#| label: fig-post-population-distribution-average

post_populDist_average <- tidybayes::gather_draws(
  fle_average_bayes
  , sigma_theta
) |>
  tidyr::nest(data = c(.chain, .iteration, .draw)) |>
  tidyr::pivot_wider(names_from = .variable, values_from = .value) |>
  mutate(
    mu_theta = 0        
  ) |>
  mutate(
    density = purrr::map2(                      # iteration of each density function for each pair of mu and sigma
      mu_theta, sigma_theta, 
      \(mu, sigma){
        dnorm(
          x = seq(-4, 7, length.out = 101)
          , mean = mu
          , sd = sigma
        )
      }
    )
    ,theta = list(seq(-4, 7, length.out = 101))    # corresponsing values on x for plot
  )
  
 theta_average_dist <- post_populDist_average |>
  tidyr::unnest(cols = c(density, theta)) |>
  ggplot() + 
  aes(x = theta, y = density) +
   geom_vline(xintercept = 0, color = "black", linetype = "22") +
   geom_hline(yintercept = 0, color = "black") + 
   geom_lineribbon(                              # function designed to use with output from point_interval
     aes(ymin = after_stat(ymin), ymax = after_stat(ymax))
     , stat = "summary"
     , fun.data = "median_qi"
     , color = "black"
     , fill = "black"
     , alpha = 0.2
   ) + 
   scale_fill_viridis_d(option = "B", begin = 0.9, end = 0.1) +
   guides(fill = "none") +
   labs(x = expression(theta[i]), y = "Density") + 
   papaja::theme_apa(base_size = 16)

```

```{r}
#| label: fig-posterior-theta-distribution-ave rage
#| fig-cap: "Posterior distribution of individual effects $\\theta_i$ for each participant (left) and the posterior distribution of the population distribution of individual effects $\\theta_i$ (right). The horizontal dashed line marks the point of no effect, i.e., $\\theta_i = 0$."

average_caterpillar + (theta_average_dist + coord_flip() + theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())) +
  plot_layout(ncol = 2, width = c(1, 1/3), axes = "collect_y", guides = "collect")
```


---

### Bayes factor with unconstrained model

```{r}
#| label: bayes-factor-average

average_ml <- bridge_sampler(fle_average_bayes, method = "warp3", repetitions = 30, silent = TRUE)

bayes_factor(unconstrained_ml, average_ml)

```

## (4) Null vs unconstrained

```{r}
#| label: fit-null-model

pse_null_model <- readLines("models/pse-null.stan") |>
  paste(collapse = "\n")

fle_stan_null_data <- list(
  N = nrow(FLE_data) |> as.integer()
  , n_subjects = length(unique(FLE_data$ID)) |> as.integer()
  , id = as.integer(as.factor(FLE_data$ID))
  , y = FLE_data$FLE_dva
)

fle_null_bayes <- rstan::stan(
  model_code = pse_null_model
  , data = fle_stan_null_data
  , iter = 1e5/4 + 1000
  , warmup = 1000
  , chains = 4
  , cores = 4
  , seed = 123
)

# model diagnostics
str_post_null <- data.frame(summary(fle_null_bayes)$summary)
check_hmc_diagnostics(fle_null_bayes)
```

### Bayes factor with unconstrained model

```{r}
#| label: bayes-factor-null

null_ml <- bridge_sampler(fle_null_bayes, method = "warp3", repetitions = 30, silent = TRUE)

bayes_factor(unconstrained_ml, null_ml)
```

---

# Alternative dominance models
## Log-normal distributed $\theta_i$

In this model priors are put on the mean ($\bar{x}$) and the standard deviation ($s$) of the log-normal distribution. We chose this approach because we assume that expressing prior beliefs about the mean and standard deviation of $\theta_i$ is more intuitive than expressing prior beliefs about the parameters $\mu$ and $\sigma$ of the log-normal distribution.  

```{r}
#| label: fit-logN-model

mean_prior_log <- rlnorm(1e5, meanlog = 0.5, sdlog = 1)
sd_prior_log <-rlnorm(1e5, meanlog = 0.5, sdlog = 1)

pse_logN_model <- readLines("models/pse-dominance-logN.stan") |>
  paste(collapse = "\n")

fle_stan_logN_data <- list(
  N = nrow(FLE_data) |> as.integer()
  , n_subjects = length(unique(FLE_data$ID)) |> as.integer()
  , id = as.integer(as.factor(FLE_data$ID))
  , y = FLE_data$FLE_dva
  
  # Prior settings
  , mean_logN_mu = 0.5
  , mean_logN_sigma = 1
  , sd_logN_mu = 0.5
  , sd_logN_sigma = 1
)

fle_logN_bayes <- rstan::stan(
  model_code = pse_logN_model
  , data = fle_stan_logN_data
  , iter = 1e5/4 + 1000
  , warmup = 1000
  , chains = 4
  , cores = 4
  , seed = 123
)

# model diagnostics
str_post_logN <- data.frame(summary(fle_logN_bayes)$summary)
check_hmc_diagnostics(fle_logN_bayes)
```

### Estimated of individual effects 

```{r}
#| label: fig-indiv-effects-log-normal

theta_i_logN <- tidybayes::gather_draws(
  fle_logN_bayes
  , theta[i]
)

theta_i_logN_ordered <- tidybayes::median_qi(theta_i_logN) |>
  mutate(i = factor(i
                    ,levels = unique(i)[order(.value)]        
                    ,labels = unique(i)[order(.value)]        #I create this solely for creating levels in draws, as geom-f calls on draws
                    ) 
         )

# Plot median estimates
log_normal_caterpillar <- theta_i_logN |>
  mutate(ID = factor(i
                     ,levels = unique(theta_i_logN_ordered$i)[order(theta_i_logN_ordered$.value)]  #new var ID with ordered med levels
                     ,labels = unique(theta_i_logN_ordered$i)[order(theta_i_logN_ordered$.value)]
                     )
  ) |>
  rename(theta = .value) |>
  ggplot() +
  aes(x = ID, y = theta) +
  geom_hline(yintercept = 0, color = "black", linetype = "22") + #string-linetype: line for 2 units, followed by 2 off
  stat_pointinterval(           #draws point est + CI for draw-input, point_interval function is median_qi by default
    fill = viridisLite::inferno(1, begin = 0.9), interval_color = viridisLite::inferno(1, begin = 0.9), 
    point_color = "white", .width = c(.95), fatten_point = 2, linewidth = 1, shape = 21
  ) +
  labs(y = expression(theta[i]), x = "Participant ID") +
  coord_cartesian(ylim = c(0,7)) +
  papaja::theme_apa(base_size = 16) + 
  theme(
    axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5, hjust = 0)
  )
  
```

```{r}
#| label: fig-post-population-distribution-log-normal

post_populDist_logN <- tidybayes::gather_draws(
  fle_logN_bayes
  , mu_theta, sigma_theta
) |>  
  tidyr::nest(data = c(.chain, .iteration, .draw)) |>
  tidyr::pivot_wider(names_from = .variable, values_from = .value) |>
  mutate(
    density = purrr::map2(                      # iteration of each density function for each pair of mu and sigma
      mu_theta, sigma_theta, 
      \(mu, sigma){
        dlnorm(
          x = seq(0, 7, length.out = 101)
          , meanlog = mu
          , sdlog = sigma
        )
      }
    )
    ,theta = list(seq(0, 7, length.out = 101))    # corresponsing values on x for plot
  )
  
 theta_logN_dist <- post_populDist_logN |>
  tidyr::unnest(cols = c(density, theta)) |>
  ggplot() + 
  aes(x = theta, y = density) +
   geom_vline(xintercept = 0, color = "black", linetype = "22") +
   geom_hline(yintercept = 0, color = "black") + 
   geom_lineribbon(                              # function designed to use with output from point_interval
     aes(ymin = after_stat(ymin), ymax = after_stat(ymax))
     , stat = "summary"
     , fun.data = "median_qi"
     , color = "black"
     , fill = "black"
     , alpha = 0.2
   ) + 
   labs(x = expression(theta[i]), y = "Density") + 
   papaja::theme_apa(base_size = 16)

```

```{r}
#| label: fig-posterior-theta-distribution-log-normal
#| fig-cap: "Posterior distribution of individual effects $\\theta_i$ for each participant (left) and the posterior distribution of the population distribution of individual effects $\\theta_i$ (right). The horizontal dashed line marks the point of no effect, i.e., $\\theta_i = 0$."

log_normal_caterpillar + (theta_logN_dist + coord_flip() + theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())) +
  plot_layout(ncol = 2, width = c(1, 1/3), axes = "collect_y")
```

### Bayes factor with unconstrained model

```{r}
#| label: bayes-factor-logN

logN_ml <- bridge_sampler(fle_logN_bayes, method = "warp3", repetitions = 30, silent = TRUE)
bayes_factor(unconstrained_ml, logN_ml)
bayes_factor(logN_ml, dominance_ml) -> bf_logN_dom
```

---

# Prior settings
## Prior settings for mean of $\theta_i$

To assess the (possible) impact of different prior assumptions on the results of the model comparison, the prior distributions for the mean of $theta_i$ are plotted below. 
Since the common model assumes one fixed effect shared by all individuals, the prior on the common effect can be understood as a prior on the mean effect. 

```{r}
#| label: prior-mean-cross-model-comparison
#| fig-cap: "Comparison of priors for $\\mu_{\\theta}$ across the unconstrained, log-normal and common model (truncated tbd)"

# Calculate prior implied for mean of theta_i of truncated normal model
trunc_alpha <- (0-mu_prior_truncn)/sigma_prior_truncn
trunc_Z <- 1-pnorm(trunc_alpha)
mean_prior_truncn <- mu_prior_truncn + sigma_prior_truncn*(dnorm(trunc_alpha)/trunc_Z)

# Plot
data.frame(priors = c(mu_prior
                      ,mean_prior_log
                      ,mean_prior_truncn
                      ,mu_prior_common)
                      ,model = rep(c("unconstrained","log-normal","truncated normal","common"), each = 1e5)) |>
  ggplot() +
  aes(x = priors, fill = model) +
  lims(x = c(0,5)) +
  geom_density(alpha = 0.2, linewidth = 0.5) +
  labs(
    x = expression(mu[theta])
    , y = "Density"
    , fill = "Prior for"
  ) +
  scale_color_viridis_d(option = "B", begin = 0.1, end = 0.8, aesthetics = "fill") +
  papaja::theme_apa(base_size = 16)
   
```

## Prior settings for standard deviation of $\theta_i$

To assess the (possible) impact of different prior assumptions on the results of the model comparison, the prior distributions for the standard deviation of $theta_i$ are plotted below. 


```{r}
#| label: prior-sd-cross-model-comparison
#| fig-cap: "Comparison of priors for $\\sigma_{\\theta}$ across the unconstrained, log-normal and average null model (truncated tbd)"

# Calculate implied prior for sd of theta_i of truncated normal model
sd_prior_truncn <- sqrt(
  (sigma_prior_truncn^2)*
    (1 + trunc_alpha*dnorm(trunc_alpha)/trunc_Z -(dnorm(trunc_alpha)/trunc_Z)^2)
)

# Plot
data.frame(priors = c(sigma_prior
                      ,sd_prior_log
                      ,sd_prior_truncn
                      ,sigma_prior_average)
                      ,model = rep(c("unconstrained","log-normal","truncated normal","average null"), each = 1e5)) |>
  ggplot() +
  aes(x = priors, fill = model) +
  lims(x = c(0,5)) +
  geom_density(alpha = 0.2, linewidth = 0.5) +
  labs(
    x = expression(sigma[theta])
    , y = "Density"
    , fill = "Prior for"
  ) +
  scale_color_viridis_d(option = "G", begin = 0.1, end = 0.8, aesthetics = "fill") +
  papaja::theme_apa(base_size = 16)
   
```

# Bayesian model comparison

## Bayes factor summary table

```{r}
#| label: bf-summary-table
#| tbl-cap: "Summary table: Bayes Factors over unconstrained model (=reference model) and naive posterior probability (NPP), based on assumption of equal prior model odds. Maximum $\\hat{R}$ and coefficient of variation (CV) for the bridge sampling are also given for each model."

bfs_base_unc <- c(bayes_factor(null_ml, unconstrained_ml)$bf_median_based
    , bayes_factor(average_ml, unconstrained_ml)$bf_median_based
    , bayes_factor(common_ml, unconstrained_ml)$bf_median_based
    , bayes_factor(logN_ml, unconstrained_ml)$bf_median_based
    , bayes_factor(dominance_ml, unconstrained_ml)$bf_median_based
    , 1
    )

# Naive posterior probability: naive assumption that prior odds = 1 for all model comparisons
# -> post prob = bf / sum(bf)

npp_base_unc <- bfs_base_unc / sum(bfs_base_unc) 

# model diagnostics
rhat <- c(max(str_post_null$Rhat)
          ,max(str_post_average$Rhat)
          ,max(str_post_common$Rhat)
          ,max(str_post_logN$Rhat)
          ,max(str_post_dominance$Rhat)
          ,max(str_post_unconstrained$Rhat))


# bridge sampling CV
 bridge.cv <- c(sd(null_ml$logml)/abs(mean(null_ml$logml))
                ,sd(average_ml$logml)/abs(mean(average_ml$logml))
                ,sd(common_ml$logml)/abs(mean(common_ml$logml))
                ,sd(logN_ml$logml)/abs(mean(logN_ml$logml))
                ,sd(dominance_ml$logml)/abs(mean(dominance_ml$logml))
                ,sd(unconstrained_ml$logml)/abs(mean(unconstrained_ml$logml)))

# i was getting error for `printnum()` function, so asked chatgpt to define a function

printnum <- function(
    x,
    format = "f",
    digits = 3,
    gt1 = TRUE
) {

  formatted <- formatC(
    x,
    format = format,
    digits = digits
  )

  # Optional formatting for values greater than 1
  if (!gt1) {
    formatted[x > 1] <- formatC(
      x[x > 1],
      format = format,
      digits = digits
    )
  }

  formatted
}

# summary table
data.frame(
  model = c("Null model", "Average null model", "Common model", "Log-normal model", "Truncated normal model", "Unconstrained model")
  , bf = printnum(bfs_base_unc, format = "e")
  , npp = printnum(npp_base_unc, gt1 = F, digits = 2)
  , rhat = printnum(rhat, digits = 3)
  , cv = printnum(bridge.cv, format = "e")
)  |> knitr::kable(
  col.names = c("Model ($i$)", "$BF_{i/unconstrained}$", "NPP", "max.$\\hat{R}$", "Bridge Sampling CV")
)

```

---

# Summary


```{r}
#| label: bf-model-averaging

bf_dom_null <- c(bayes_factor(logN_ml,null_ml)$bf_median_based
                 ,bayes_factor(dominance_ml, null_ml)$bf_median_based)
bf_unc_null <- bayes_factor(unconstrained_ml, null_ml)$bf_median_based

av_bf_dom_unc <- length(bf_dom_null)^(-1)*(sum(bf_dom_null)) / bf_unc_null

```

