---
title: "Data Cleaning New Experiment"
date: today
author: "Imane Lattab"
institute: "The University of Sydney"
repository: https://github.com/imanel04/PsychHonours.git
format: 
  html:
    embed-resources: true
    code-fold: true
    code-tools: true
    theme: sandstone
    fig_caption: yes
    table-of-contents: true
    toc: true
    toc_depth: 4
    toc_float: true
execute:
  echo: true
  tidy: true
number-sections: false 
---
```{r setup, include=FALSE}
# LIBRARIES
library(R.matlab)
library(quid)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggExtra)
library(patchwork)
library(purrr)

# from previous HM report i worked on
library(viridis)
library(plotly)
library(DT)
library(lme4)
library(lmerTest) 
library(ez)
library(broom.mixed)

# MIXED MODELS
library(lme4) # mle
library(lmerTest)
library(emmeans)
library(parameters)
library(performance)

# BAYESIAN
library(rstan)
library(tidybayes)
library(logspline) # density estimation
library(bridgesampling) # bayes factors
library(brms)

# for downloading OSF repos
library(osfr)
```

# Raw Data Pre-processing

## Data cleaning


::: panel-tabset

### Read in raw data

```{r}
# raw data folder path (N=88)
data_folder <- "/Users/imanelattab/Desktop/Honours/Data/Raw_Data_FULL" # ADJUST TO YOUR OWN PATH!

# import participant fikes
file_paths <- list.files(
  path = data_folder,
  pattern = "\\.csv$",
  full.names = TRUE
)

raw_FLE_data <- map_dfr(file_paths, \(file_path) {
  
  read_csv(
    file_path,
    na = c("", "NA", "NaN"),
    show_col_types = FALSE
  ) |>
    mutate(
      # extract IDs such as "001" from the beginning of each filename
      ID = str_extract(basename(file_path), "^\\d+"),
      source_file = basename(file_path),
      source_row = row_number(),
      .before = 1
    )
})

# Participants in the raw data
n_distinct(raw_FLE_data$ID)

# Cross-check against the number of CSV files (should match, and be 88)
length(file_paths)

# Rows per participant (spots participants with unusually few/many rows)
raw_FLE_data |>
  count(ID, name = "raw_rows") |>
  arrange(raw_rows)
```

### Exclusion criteria

```{r}
# <---------EXCLUSION CRITERIA--------->
FLE_data_flagged <- raw_FLE_data |>
  mutate(
    missing_foveal_offset = is.na(foveal_offset),
    
    excessive_skipped_frames =
      !is.na(stim_skipped_frames) &
      stim_skipped_frames > 1,
    
    exclusion_reason = case_when(
      missing_foveal_offset & excessive_skipped_frames ~
        "Missing foveal offset and >1 skipped frames",
      
      missing_foveal_offset ~
        "Missing foveal offset",
      
      excessive_skipped_frames ~
        ">1 skipped frames",
      
      TRUE ~ NA_character_
    ),
    
    excluded = !is.na(exclusion_reason)
  )
```

### Clean block data

```{r}
# <--------new cleaned data---------->
clean_FLE_trials <- FLE_data_flagged |>
  filter(!excluded) |>
  mutate(
    # retain the original signed offset
    foveal_offset_og = foveal_offset,
    
    # reverse sign for only petal trials
    FLE_dva = if_else(
      motion_direction == "petal",
      -foveal_offset,
      foveal_offset
    )
  ) |>
  select(
    ID,
    source_file,
    block,
    trial,
    condition,
    motion_direction,
    eccentricity_dva,
    speed_dva,
    foveal_offset_og,
    FLE_dva,
    everything(),
    -foveal_offset,
    -excluded,
    -exclusion_reason,
    -missing_foveal_offset,
    -excessive_skipped_frames
  ) |>
  arrange(ID, block, trial)
```

### Checking for valid blocks & trials

Here, I exclude participant 052 for having completed only 1 block.

I also flag participants with <75% "valid" blocks,  where a valid block is defined as having ≥75% valid trials. I will exclude these participants from analyses and see if this changes results, then discuss this slightly poor data quality as a limitation if so.


```{r}
# <--------BLOCK QUALITY AND PARTICIPANT EXCLUSIONS------->
min_valid_prop <- 0.75  # criterion for flagging a block
min_valid_blocks <- 9   # criterion for flagging a participant
min_blocks_to_keep <- 2 # exclude only participants with 0 or 1 block

# Calculate trial validity within every available block
block_quality <- FLE_data_flagged |>
  filter(
    condition == "flashbaseline" |
      (
        condition == "centralcue" &
          motion_direction %in% c("petal", "fugal")
      )
  ) |>
  group_by(ID, block) |>
  summarise(
    n_total = n(),
    n_valid = sum(!excluded),
    prop_valid = n_valid / n_total,
    .groups = "drop"
  ) |>
  mutate(
    # This is now a flag only: blocks are not removed
    block_ok = prop_valid >= min_valid_prop,
    block_below_75 = prop_valid < min_valid_prop
  )
```

```{r}
participant_blocks <- block_quality |>
  group_by(ID) |>
  summarise(
    # Number of blocks the participant completed
    n_blocks_total = n_distinct(block),
    
    # Number satisfying the 75% valid-trial rule
    n_blocks_valid = sum(block_ok, na.rm = TRUE),
    
    .groups = "drop"
  ) |>
  mutate(
    # Only participants with one or fewer completed blocks are excluded
    participant_ok = n_blocks_total >= min_blocks_to_keep,
    
    # Flag participants with fewer than 9 valid blocks
    fewer_than_9_valid_blocks = n_blocks_valid < min_valid_blocks,
    
    # Optional more descriptive label
    valid_block_status = if_else(
      fewer_than_9_valid_blocks,
      "<9 valid blocks",
      "≥9 valid blocks"
    )
  )
```

```{r}
# Save the trial-level dataset before participant/block filtering
clean_FLE_trials_all <- clean_FLE_trials

clean_FLE_trials <- clean_FLE_trials_all |>
  # Remove only participants with 0 or 1 completed block
  semi_join(
    participant_blocks |>
      filter(participant_ok),
    by = "ID"
  ) |>
  
  # Attach block-level validity information
  left_join(
    block_quality |>
      select(
        ID,
        block,
        block_n_total = n_total,
        block_n_valid = n_valid,
        block_prop_valid = prop_valid,
        block_ok,
        block_below_75
      ),
    by = c("ID", "block")
  ) |>
  
  # Attach participant-level validity information
  left_join(
    participant_blocks |>
      select(
        ID,
        n_blocks_total,
        n_blocks_valid,
        fewer_than_9_valid_blocks,
        valid_block_status
      ),
    by = "ID"
  )
```


### Inspect blocks & retained IDs

```{r}
cat(
  "Blocks flagged as <75% valid:",
  sum(block_quality$block_below_75),
  "of",
  nrow(block_quality),
  "\n"
)

cat(
  "Participants flagged as having <9 valid blocks:",
  sum(participant_blocks$fewer_than_9_valid_blocks),
  "of",
  nrow(participant_blocks),
  "\n"
)

cat(
  "Participants excluded because they completed <=1 block:",
  sum(!participant_blocks$participant_ok),
  "\n"
)

cat(
  "Participants retained:",
  sum(participant_blocks$participant_ok),
  "\n"
)
```
```{r}
# Blocks below 75% validity—retained but flagged
block_quality |>
  filter(block_below_75) |>
  arrange(ID, block)

# Participants with fewer than nine valid blocks—retained if they completed >1 block
participant_blocks |>
  filter(fewer_than_9_valid_blocks) |>
  arrange(n_blocks_valid)
```


:::

## Collapsing over speed and eccentricity

This will leave us with motion direction as the only condition. I will be using this data analysis for the initial analysis on whether not qualitative individual differences exist (comparing "fugal" and "petal" foveal motion).

::: panel-tabset

### Average over conditions

```{r}

# <----------aggregate over conditions per block---------->

FLE_block_data <- clean_FLE_trials |>
  mutate(
    # Convert each individual motion trial from dva to ms
    FLE_ms_trial = if_else(
      condition == "centralcue" & !is.na(speed_dva),
      (FLE_dva / speed_dva) * 1000,
      NA_real_
    ),
    
    FLE_condition = case_when(
      condition == "flashbaseline" ~ "flashbaseline",
      
      condition == "centralcue" &
        motion_direction == "petal" ~ "petal",
      
      condition == "centralcue" &
        motion_direction == "fugal" ~ "fugal",
      
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(FLE_condition)) |>
  group_by(
    ID,
    block,
    FLE_condition
  ) |>
  summarise(
    # Spatial average across speed and eccentricity
    FLE_dva = mean(FLE_dva),
    
    # Temporal average after trial-level conversion
    FLE_ms = if_else(
      all(is.na(FLE_ms_trial)),
      NA_real_,
      mean(FLE_ms_trial, na.rm = TRUE)
    ),
    
    n_trials = n(),
    n_temporal_trials = sum(!is.na(FLE_ms_trial)),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("flashbaseline", "petal", "fugal")
    )
  ) |>
  arrange(
    ID,
    block,
    FLE_condition
  )

```

### Check Block Data

```{r}
# <-------check number of retained rows-------->
FLE_block_check <- FLE_block_data |>
  mutate(
    expected_trials = case_when(
      FLE_condition == "petal" ~ 12L,
      FLE_condition == "fugal" ~ 12L,
      FLE_condition == "flashbaseline" ~ 8L
    ),
    missing_trials = expected_trials - n_trials,
    complete_cell = n_trials == expected_trials
  )

FLE_block_check
```

```{r}
#<--------exclusion report-------->
excluded_FLE_trials <- FLE_data_flagged |>
  filter(excluded) |>
  select(
    ID,
    source_file,
    source_row,
    block,
    trial,
    condition,
    motion_direction,
    eccentricity_dva,
    speed_dva,
    foveal_offset,
    stim_skipped_frames,
    exclusion_reason
  ) |>
  arrange(ID, block, trial)

excluded_FLE_trials
```

```{r}
# <----------participant level summary--------->
exclusion_summary <- FLE_data_flagged |>
  group_by(ID) |>
  summarise(
    raw_rows = n(),
    missing_foveal_offset = sum(missing_foveal_offset),
    skipped_frames_gt_1 = sum(excessive_skipped_frames),
    total_excluded = sum(excluded),
    retained_trials = sum(!excluded),
    .groups = "drop"
  )
```

:::

## Preserving speed

::: panel-tabset

### Average over eccentricity

```{r}
# <------average over eccentricity, but retain speed------->

FLE_block_speed_data <- clean_FLE_trials |>
  mutate(
    FLE_condition = case_when(
      condition == "flashbaseline" ~ "flashbaseline",
      
      condition == "centralcue" &
        motion_direction == "petal" ~ "petal",
      
      condition == "centralcue" &
        motion_direction == "fugal" ~ "fugal",
      
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(FLE_condition)) |>
  group_by(
    ID,
    block,
    FLE_condition,
    speed_dva
  ) |>
  summarise(
    FLE_dva = mean(FLE_dva),
    n_trials = n(),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("flashbaseline", "petal", "fugal")
    )
  ) |>
  arrange(
    ID,
    block,
    FLE_condition,
    speed_dva
  )

# <----FLE in temporal units----->
FLE_block_speed_data <- FLE_block_speed_data |>
  mutate(
    FLE_ms = if_else(
      FLE_condition %in% c("petal", "fugal") &
        !is.na(speed_dva),
      (FLE_dva / speed_dva) * 1000,
      NA_real_
    )
  )
```

### Check Speed Data

```{r}
# <--check number of retained trials per aggregated row------>

FLE_block_speed_check <- FLE_block_speed_data |>
  mutate(
    expected_trials = case_when(
      FLE_condition %in% c("petal", "fugal") ~ 4L,
      FLE_condition == "flashbaseline" ~ 8L,
      TRUE ~ NA_integer_
    ),
    
    missing_trials = expected_trials - n_trials,
    complete_cell = n_trials == expected_trials
  )

FLE_block_speed_check
```

:::

## Retaining all trials and conditions


```{r}
# <------trial-level data: no aggregation------->
FLE_trial_data <- clean_FLE_trials |>
  mutate(
    FLE_condition = case_when(
      condition == "flashbaseline" ~ "flashbaseline",
      condition == "centralcue" & motion_direction == "petal" ~ "petal",
      condition == "centralcue" & motion_direction == "fugal" ~ "fugal",
      TRUE ~ NA_character_
    ),
    # temporal units, converted per trial (NA for flash baseline)
    FLE_ms = if_else(
      FLE_condition %in% c("petal", "fugal") & !is.na(speed_dva),
      (FLE_dva / speed_dva) * 1000,
      NA_real_
    )
  ) |>
  filter(!is.na(FLE_condition)) |>
  mutate(
    FLE_condition = factor(FLE_condition,
                           levels = c("flashbaseline", "petal", "fugal"))
  ) |>
  select(
    ID, block, trial,
    FLE_condition, motion_direction,
    eccentricity_dva, speed_dva,
    foveal_offset_og, FLE_dva, FLE_ms,
    stim_skipped_frames
  ) |>
  arrange(ID, block, trial)
```

---

## Write and Read in cleaned data files

```{r}
write_csv(FLE_trial_data, "data/FLE_Sep2026_trial_data.csv")
FLE_block_data <- read.csv("data/FLE_Sep2026_block_data.csv")

write_csv(FLE_trial_data, "data/FLE_Sep2026_block_speed_data.csv")
FLE_block_speed_data <- read.csv("data/FLE_Sep2026_block_speed_data.csv")

write_csv(FLE_trial_data, "data/FLE_Sep2026_trial_data.csv")
FLE_trial_data <- read.csv("data/FLE_Sep2026_trial_data.csv")
```