---
title: "NewExperimentData"
date: today
author: "Imane Lattab"
institute: "The University of Sydney"
repository: https://github.com/imanel04/PsychHonours.git
format: 
  html:
    embed-resources: true
    code-fold: true
    code-tools: true
    theme: sandstone
    fig_caption: yes
    table-of-contents: true
    toc: true
    toc_depth: 4
    toc_float: true
execute:
  echo: true
  tidy: true
number-sections: false 
---

```{r setup, include=FALSE}
# LIBRARIES
library(R.matlab)
library(quid)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggExtra)
library(patchwork)

# from previous HM report i worked on
library(viridis)
library(plotly)
library(DT)
library(lme4)
library(lmerTest) 
library(ez)
library(broom.mixed)

# MIXED MODELS
library(lme4) # mle
library(lmerTest)
library(emmeans)
library(performance)

# BAYESIAN
library(rstan)
library(tidybayes)
library(logspline) # density estimation
library(bridgesampling) # bayes factors
library(brms)

# for downloading OSF repos
library(osfr)
```

# Data Pre-processing

::: panel-tabset

## Collapsing over speed and eccentricity

This will leave us with motion direction as the only condition. I will be using this data analysis for the initial analysis on whether not qualitative individual differences exist (comparing "fugal" and "petal" foveal motion).

```{r}
# raw data folder path (N=88)
data_folder <- "/Users/imanelattab/Desktop/Honours/Data/Raw_Data_FULL" # ADJUST TO YOUR OWN PATH!

# import participant fikes
file_paths <- list.files(
  path = data_folder,
  pattern = "\\.csv$",
  full.names = TRUE
)

raw_FLE_data <- map_dfr(file_paths, \(file_path) {
  
  read_csv(
    file_path,
    na = c("", "NA", "NaN"),
    show_col_types = FALSE
  ) |>
    mutate(
      # extract IDs such as "001" from the beginning of each filename
      ID = str_extract(basename(file_path), "^\\d+"),
      source_file = basename(file_path),
      source_row = row_number(),
      .before = 1
    )
})
```

```{r}

# <---------EXCLUSION CRITERIA--------->
FLE_data_flagged <- raw_FLE_data |>
  mutate(
    missing_foveal_offset = is.na(foveal_offset),
    
    excessive_skipped_frames =
      !is.na(stim_skipped_frames) &
      stim_skipped_frames > 1,
    
    exclusion_reason = case_when(
      missing_foveal_offset & excessive_skipped_frames ~
        "Missing foveal offset and >1 skipped frames",
      
      missing_foveal_offset ~
        "Missing foveal offset",
      
      excessive_skipped_frames ~
        ">1 skipped frames",
      
      TRUE ~ NA_character_
    ),
    
    excluded = !is.na(exclusion_reason)
  )
```

```{r}
# <--------new cleaned data---------->
clean_FLE_trials <- FLE_data_flagged |>
  filter(!excluded) |>
  mutate(
    # retain the original signed offset
    foveal_offset_og = foveal_offset,
    
    # reverse sign for only petal trials
    FLE_dva = if_else(
      motion_direction == "petal",
      -foveal_offset,
      foveal_offset
    )
  ) |>
  select(
    ID,
    source_file,
    block,
    trial,
    condition,
    motion_direction,
    eccentricity_dva,
    speed_dva,
    foveal_offset_og,
    FLE_dva,
    everything(),
    -foveal_offset,
    -excluded,
    -exclusion_reason,
    -missing_foveal_offset,
    -excessive_skipped_frames
  ) |>
  arrange(ID, block, trial)
```

```{r}

# <----------aggregate over conditions per block---------->

FLE_block_data <- clean_FLE_trials |>
  mutate(
    # Convert each individual motion trial from dva to ms
    FLE_ms_trial = if_else(
      condition == "centralcue" & !is.na(speed_dva),
      (FLE_dva / speed_dva) * 1000,
      NA_real_
    ),
    
    FLE_condition = case_when(
      condition == "flashbaseline" ~ "flashbaseline",
      
      condition == "centralcue" &
        motion_direction == "petal" ~ "petal",
      
      condition == "centralcue" &
        motion_direction == "fugal" ~ "fugal",
      
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(FLE_condition)) |>
  group_by(
    ID,
    block,
    FLE_condition
  ) |>
  summarise(
    # Spatial average across speed and eccentricity
    FLE_dva = mean(FLE_dva),
    
    # Temporal average after trial-level conversion
    FLE_ms = if_else(
      all(is.na(FLE_ms_trial)),
      NA_real_,
      mean(FLE_ms_trial, na.rm = TRUE)
    ),
    
    n_trials = n(),
    n_temporal_trials = sum(!is.na(FLE_ms_trial)),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("flashbaseline", "petal", "fugal")
    )
  ) |>
  arrange(
    ID,
    block,
    FLE_condition
  )

```

```{r}
# <-------check number of retained rows-------->
FLE_block_check <- FLE_block_data |>
  mutate(
    expected_trials = case_when(
      FLE_condition == "petal" ~ 12L,
      FLE_condition == "fugal" ~ 12L,
      FLE_condition == "flashbaseline" ~ 8L
    ),
    missing_trials = expected_trials - n_trials,
    complete_cell = n_trials == expected_trials
  )

FLE_block_check
```

```{r}
#<--------exclusion report-------->
excluded_FLE_trials <- FLE_data_flagged |>
  filter(excluded) |>
  select(
    ID,
    source_file,
    source_row,
    block,
    trial,
    condition,
    motion_direction,
    eccentricity_dva,
    speed_dva,
    foveal_offset,
    stim_skipped_frames,
    exclusion_reason
  ) |>
  arrange(ID, block, trial)

excluded_FLE_trials
```

```{r}
# <----------participant level summary--------->
exclusion_summary <- FLE_data_flagged |>
  group_by(ID) |>
  summarise(
    raw_rows = n(),
    missing_foveal_offset = sum(missing_foveal_offset),
    skipped_frames_gt_1 = sum(excessive_skipped_frames),
    total_excluded = sum(excluded),
    retained_trials = sum(!excluded),
    .groups = "drop"
  )
```

```{r}
#<-------------WRITE NEW CSVs--------->
write_csv(FLE_block_check, "FLE_Sep2026_block_data.csv")
```

## Preserving speed

```{r}
# <------aggregate over eccentricity, but retain speed------->

FLE_block_speed_data <- clean_FLE_trials |>
  mutate(
    FLE_condition = case_when(
      condition == "flashbaseline" ~ "flashbaseline",
      
      condition == "centralcue" &
        motion_direction == "petal" ~ "petal",
      
      condition == "centralcue" &
        motion_direction == "fugal" ~ "fugal",
      
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(FLE_condition)) |>
  group_by(
    ID,
    block,
    FLE_condition,
    speed_dva
  ) |>
  summarise(
    FLE_dva = mean(FLE_dva),
    n_trials = n(),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("flashbaseline", "petal", "fugal")
    )
  ) |>
  arrange(
    ID,
    block,
    FLE_condition,
    speed_dva
  )

# <----FLE in temporal units----->
FLE_block_speed_data <- FLE_block_speed_data |>
  mutate(
    FLE_ms = if_else(
      FLE_condition %in% c("petal", "fugal") &
        !is.na(speed_dva),
      (FLE_dva / speed_dva) * 1000,
      NA_real_
    )
  )
```

```{r}
# <--check number of retained trials per aggregated row------>

FLE_block_speed_check <- FLE_block_speed_data |>
  mutate(
    expected_trials = case_when(
      FLE_condition %in% c("petal", "fugal") ~ 4L,
      FLE_condition == "flashbaseline" ~ 8L,
      TRUE ~ NA_integer_
    ),
    
    missing_trials = expected_trials - n_trials,
    complete_cell = n_trials == expected_trials
  )

FLE_block_speed_check
```

```{r}
write_csv(
  FLE_block_speed_check,
  "FLE_Sep2026_block_speed_data.csv"
)
```



:::

# Data Visualisations

::: panel-tabset

## Histogram (DVA)

```{r}
FLE_motion_data <- FLE_block_data |>
  filter(FLE_condition %in% c("petal", "fugal")) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("petal", "fugal"),
      labels = c("Petal", "Fugal")
    )
  )

FLE_participant_means <- FLE_motion_data |>
  group_by(ID, FLE_condition) |>
  summarise(
    mean_FLE = mean(FLE_dva, na.rm = TRUE),
    n_blocks = sum(!is.na(FLE_dva)),
    .groups = "drop"
  )

FLE_histogram_stats <- FLE_participant_means |>
  group_by(FLE_condition) |>
  summarise(
    n = sum(!is.na(mean_FLE)),
    mean = mean(mean_FLE, na.rm = TRUE),
    sd = sd(mean_FLE, na.rm = TRUE),
    variance = var(mean_FLE, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    lower_sd = mean - sd,
    upper_sd = mean + sd,
    statistics_label = paste0(
      "N = ", n,
      "\nMean = ", round(mean, 2), " dva",
      "\nSD = ", round(sd, 2), " dva",
      "\nVariance = ", round(variance, 2), " dva²"
    )
  )
```

```{r}
# PLOT
ggplot(
  FLE_participant_means,
  aes(x = mean_FLE, fill = FLE_condition)
) +
  geom_histogram(
    bins = 20,
    colour = "white",
    linewidth = 0.4,
    show.legend = FALSE
  ) +
  
  # Zero
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 1
  ) +
  
  # Condition mean
  geom_vline(
    data = FLE_histogram_stats,
    aes(xintercept = mean),
    colour = "darkgreen",
    linetype = "dashed",
    linewidth = 1
  ) +
  
  # Mean - 1 SD
  geom_vline(
    data = FLE_histogram_stats,
    aes(xintercept = lower_sd),
    colour = "black",
    linetype = "dotted",
    linewidth = 0.8
  ) +
  
  # Mean + 1 SD
  geom_vline(
    data = FLE_histogram_stats,
    aes(xintercept = upper_sd),
    colour = "black",
    linetype = "dotted",
    linewidth = 0.8
  ) +
  
  geom_label(
    data = FLE_histogram_stats,
    aes(
      x = Inf,
      y = Inf,
      label = statistics_label
    ),
    hjust = 1.05,
    vjust = 1.1,
    size = 3.4,
    label.size = 0.2,
    fill = "white",
  ) +
  
  facet_wrap(~ FLE_condition, nrow = 1) +
  
  scale_fill_manual(
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  
  labs(
    title = "Distributions of participant-level flash-lag effects",
    subtitle = "Participant means across repeated blocks",
    x = "Mean FLE (dva)",
    y = "Number of participants"
  ) +
  
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 13)
  )
```

## Histogram (temporal units)

```{r}
FLE_motion_data <- FLE_block_data |>
  filter(FLE_condition %in% c("petal", "fugal")) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("petal", "fugal"),
      labels = c("Petal", "Fugal")
    )
  )

# One temporal FLE mean per participant and motion direction
FLE_participant_means_ms <- FLE_motion_data |>
  group_by(ID, FLE_condition) |>
  summarise(
    mean_FLE_ms = mean(FLE_ms, na.rm = TRUE),
    n_blocks = sum(!is.na(FLE_ms)),
    .groups = "drop"
  )

FLE_histogram_stats_ms <- FLE_participant_means_ms |>
  group_by(FLE_condition) |>
  summarise(
    n = sum(!is.na(mean_FLE_ms)),
    mean = mean(mean_FLE_ms, na.rm = TRUE),
    sd = sd(mean_FLE_ms, na.rm = TRUE),
    variance = var(mean_FLE_ms, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    lower_sd = mean - sd,
    upper_sd = mean + sd,
    
    statistics_label = paste0(
      "N = ", n,
      "\nMean = ", round(mean, 2), " ms",
      "\nSD = ", round(sd, 2), " ms",
      "\nVariance = ", round(variance, 2), " ms²"
    )
  )
```

```{r}
# PLOT
ggplot(
  FLE_participant_means_ms,
  aes(
    x = mean_FLE_ms,
    fill = FLE_condition
  )
) +
  geom_histogram(
    bins = 20,
    colour = "white",
    linewidth = 0.4,
    show.legend = FALSE
  ) +
  
  # Zero
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 1
  ) +
  
  # Condition mean
  geom_vline(
    data = FLE_histogram_stats_ms,
    aes(xintercept = mean),
    colour = "darkgreen",
    linetype = "dashed",
    linewidth = 1,
    inherit.aes = FALSE
  ) +
  
  # Mean - 1 SD
  geom_vline(
    data = FLE_histogram_stats_ms,
    aes(xintercept = lower_sd),
    colour = "black",
    linetype = "dotted",
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  
  # Mean + 1 SD
  geom_vline(
    data = FLE_histogram_stats_ms,
    aes(xintercept = upper_sd),
    colour = "black",
    linetype = "dotted",
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  
  geom_label(
    data = FLE_histogram_stats_ms,
    aes(
      x = Inf,
      y = Inf,
      label = statistics_label
    ),
    hjust = 1.05,
    vjust = 1.1,
    size = 3.4,
    label.size = 0.2,
    fill = "white",
    inherit.aes = FALSE
  ) +
  
  facet_wrap(
    ~ FLE_condition,
    nrow = 1
  ) +
  
  scale_fill_manual(
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  
  labs(
    title = "Distributions of participant-level flash-lag effects",
    subtitle = "Participant means across repeated blocks",
    x = "Mean FLE (ms)",
    y = "Number of participants"
  ) +
  
  theme_classic(base_size = 13) +
  
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 13)
  )
```

## FLE by participant

```{r}
FLE_individual_summary <- FLE_motion_data |>
  group_by(ID, FLE_condition) |>
  summarise(
    mean_FLE = mean(FLE_dva, na.rm = TRUE),
    sd_FLE = sd(FLE_dva, na.rm = TRUE),
    n_blocks = sum(!is.na(FLE_dva)),
    .groups = "drop"
  ) |>
  mutate(
    ID = fct_reorder(factor(ID), mean_FLE)
  )

ggplot(
  FLE_individual_summary,
  aes(
    x = ID,
    y = mean_FLE,
    colour = FLE_condition
  )
) +
  geom_hline(
    yintercept = 0,
    colour = "red",
    linewidth = 0.9
  ) +
  
  geom_errorbar(
    aes(
      ymin = mean_FLE - sd_FLE,
      ymax = mean_FLE + sd_FLE
    ),
    width = 0,
    linewidth = 0.45,
    alpha = 0.7
  ) +
  
  geom_point(
    size = 1.7
  ) +
  
  coord_flip() +
  facet_wrap(
    ~ FLE_condition,
    nrow = 1,
    scales = "free_y"
  ) +
  
  scale_colour_manual(
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    ),
    guide = "none"
  ) +
  
  labs(
    title = "Individual mean FLE and within-participant variability",
    subtitle = "Points are participant means; error bars show ±1 SD across blocks",
    x = "Participant",
    y = "FLE (dva)"
  ) +
  
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 5)
  )
```

## Averaged over motion-direction

```{r}
FLE_motion_data |>
  group_by(ID) |>
  summarise(
    mean_FLE = mean(FLE_dva, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = mean_FLE)) +
  geom_histogram(
    bins = 20,
    fill = "steelblue",
    colour = "white"
  ) +
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 1
  ) +
  labs(
    x = "Mean FLE (dva)",
    y = "Count"
  ) +
  theme_classic()
```

## Descriptive stats

```{r}
# One petal and one fugal mean per participant
FLE_petal_fugal <- FLE_block_data |>
  filter(FLE_condition %in% c("petal", "fugal")) |>
  group_by(ID, FLE_condition) |>
  summarise(
    mean_FLE = mean(FLE_dva, na.rm = TRUE),
    n_blocks = sum(!is.na(FLE_dva)),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = FLE_condition,
    values_from = c(mean_FLE, n_blocks)
  )

petal_fugal_correlation <- cor.test(
  FLE_petal_fugal$mean_FLE_petal,
  FLE_petal_fugal$mean_FLE_fugal,
  method = "pearson",
  use = "complete.obs"
)

correlation_summary <- tibble(
  n = sum(
    complete.cases(
      FLE_petal_fugal$mean_FLE_petal,
      FLE_petal_fugal$mean_FLE_fugal
    )
  ),
  r = unname(petal_fugal_correlation$estimate),
  p = petal_fugal_correlation$p.value,
  CI_lower = petal_fugal_correlation$conf.int[1],
  CI_upper = petal_fugal_correlation$conf.int[2]
)

correlation_summary
```

```{r}
FLE_paired_blocks <- FLE_block_data |>
  filter(FLE_condition %in% c("petal", "fugal")) |>
  select(ID, block, FLE_condition, FLE_dva) |>
  pivot_wider(
    names_from = FLE_condition,
    values_from = FLE_dva
  )

cor(
  FLE_paired_blocks$petal,
  FLE_paired_blocks$fugal,
  use = "complete.obs",
  method = "pearson"
)
```

```{r}
# Pair petal and fugal FLE within each participant's block
FLE_paired_blocks <- FLE_block_data |>
  filter(FLE_condition %in% c("petal", "fugal")) |>
  select(ID, block, FLE_condition, FLE_dva) |>
  pivot_wider(
    names_from = FLE_condition,
    values_from = FLE_dva
  )

FLE_individual_correlations <- FLE_paired_blocks |>
  group_by(ID) |>
  summarise(
    n_blocks = sum(complete.cases(petal, fugal)),
    
    correlation = if (
      n_blocks >= 3 &&
      sd(petal, na.rm = TRUE) > 0 &&
      sd(fugal, na.rm = TRUE) > 0
    ) {
      cor(
        petal,
        fugal,
        use = "complete.obs",
        method = "pearson"
      )
    } else {
      NA_real_
    },
    
    .groups = "drop"
  )

FLE_individual_correlations

ggplot(
  FLE_individual_correlations,
  aes(x = correlation)
) +
  geom_histogram(
    bins = 20,
    fill = "steelblue",
    colour = "white",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 0.9
  ) +
  scale_x_continuous(
    limits = c(-1, 1),
    breaks = seq(-1, 1, by = 0.25)
  ) +
  labs(
    title = "Distribution of within-participant correlations",
    subtitle = "Correlation between petal and fugal FLE across blocks",
    x = "Pearson correlation coefficient",
    y = "Number of participants"
  ) +
  theme_classic(base_size = 13)
```

```{r}
# Pair matching petal and fugal trials
FLE_paired_trials <- clean_FLE_trials |>
  filter(
    condition == "centralcue",
    motion_direction %in% c("petal", "fugal")
  ) |>
  select(
    ID,
    block,
    speed_dva,
    eccentricity_dva,
    motion_direction,
    FLE_dva
  ) |>
  pivot_wider(
    names_from = motion_direction,
    values_from = FLE_dva
  ) |>
  drop_na(petal, fugal)

FLE_individual_trial_correlations <- FLE_paired_trials |>
  group_by(ID) |>
  summarise(
    n_pairs = n(),
    
    correlation = if (
      n_pairs >= 3 &&
      sd(petal, na.rm = TRUE) > 0 &&
      sd(fugal, na.rm = TRUE) > 0
    ) {
      cor(
        petal,
        fugal,
        use = "complete.obs",
        method = "pearson"
      )
    } else {
      NA_real_
    },
    
    .groups = "drop"
  )

FLE_individual_trial_correlations

ggplot(
  FLE_individual_trial_correlations,
  aes(x = correlation)
) +
  geom_histogram(
    bins = 20,
    fill = "steelblue",
    colour = "white",
    linewidth = 0.4
  ) +
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 0.9
  ) +
  scale_x_continuous(
    limits = c(-1, 1),
    breaks = seq(-1, 1, by = 0.25)
  ) +
  labs(
    title = "Distribution of within-participant trial-level correlations",
    subtitle = paste(
      "Petal and fugal trials paired by block,",
      "speed, and eccentricity"
    ),
    x = "Petal–fugal Pearson correlation",
    y = "Number of participants"
  ) +
  theme_classic(base_size = 13)

FLE_individual_trial_correlations |>
  count(n_pairs)
```


## Speed x motion direction

```{r}
# One mean FLE per participant for each motion direction × speed
FLE_direction_speed_means <- FLE_block_speed_data |>
  filter(
    FLE_condition %in% c("petal", "fugal"),
    speed_dva %in% c(10, 15, 20)
  ) |>
  group_by(
    ID,
    FLE_condition,
    speed_dva
  ) |>
  summarise(
    mean_FLE = mean(FLE_dva, na.rm = TRUE),
    n_blocks = sum(!is.na(FLE_dva)),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("petal", "fugal"),
      labels = c("Petal", "Fugal")
    ),
    speed_dva = factor(
      speed_dva,
      levels = c(10, 15, 20),
      labels = c("10 dva/s", "15 dva/s", "20 dva/s")
    )
  )

# PLOT
ggplot(
  FLE_direction_speed_means,
  aes(
    x = mean_FLE,
    fill = FLE_condition
  )
) +
  geom_histogram(
    bins = 20,
    colour = "white",
    linewidth = 0.4,
    show.legend = FALSE
  ) +
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 0.9
  ) +
  facet_grid(
    rows = vars(FLE_condition),
    cols = vars(speed_dva)
  ) +
  scale_fill_manual(
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  labs(
    title = "Participant-level FLE distributions by motion direction and speed",
    subtitle = "Each observation is one participant mean across blocks and eccentricities",
    x = "Mean FLE (dva)",
    y = "Count"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

```{r}
# One mean temporal FLE per participant × direction × speed
FLE_direction_speed_means_ms <- FLE_block_speed_data |>
  filter(
    FLE_condition %in% c("petal", "fugal"),
    speed_dva %in% c(10, 15, 20),
    !is.na(FLE_ms)
  ) |>
  group_by(
    ID,
    FLE_condition,
    speed_dva
  ) |>
  summarise(
    mean_FLE_ms = mean(FLE_ms, na.rm = TRUE),
    n_blocks = sum(!is.na(FLE_ms)),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("petal", "fugal"),
      labels = c("Petal", "Fugal")
    ),
    speed_dva = factor(
      speed_dva,
      levels = c(10, 15, 20),
      labels = c("10 dva/s", "15 dva/s", "20 dva/s")
    )
  )
```
```{r}
ggplot(
  FLE_direction_speed_means_ms,
  aes(
    x = mean_FLE_ms,
    fill = FLE_condition
  )
) +
  geom_histogram(
    bins = 20,
    colour = "white",
    linewidth = 0.4,
    show.legend = FALSE
  ) +
  geom_vline(
    xintercept = 0,
    colour = "red",
    linewidth = 0.9
  ) +
  facet_grid(
    rows = vars(FLE_condition),
    cols = vars(speed_dva)
  ) +
  scale_fill_manual(
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  labs(
    title = "Participant-level temporal FLE distributions",
    subtitle = paste(
      "Distributions by motion direction and speed;",
      "each observation is one participant mean across blocks and eccentricities"
    ),
    x = "Mean FLE (ms)",
    y = "Count"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

```{r}
# One spatial and temporal FLE per participant × speed
FLE_participant_speed_means <- FLE_block_speed_data |>
  filter(
    FLE_condition %in% c("petal", "fugal"),
    speed_dva %in% c(10, 15, 20)
  ) |>
  group_by(ID, speed_dva) |>
  summarise(
    participant_FLE_dva = mean(FLE_dva, na.rm = TRUE),
    participant_FLE_ms  = mean(FLE_ms, na.rm = TRUE),
    .groups = "drop"
  )

FLE_direction_speed_population_stats <- FLE_block_speed_data |>
  filter(
    FLE_condition %in% c("petal", "fugal"),
    speed_dva %in% c(10, 15, 20)
  ) |>
  group_by(
    ID,
    FLE_condition,
    speed_dva
  ) |>
  summarise(
    participant_FLE_dva = mean(FLE_dva, na.rm = TRUE),
    participant_FLE_ms = mean(FLE_ms, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(
    FLE_condition,
    speed_dva
  ) |>
  summarise(
    n = sum(
      complete.cases(participant_FLE_dva, participant_FLE_ms)
    ),
    
    mean_FLE_dva = mean(participant_FLE_dva, na.rm = TRUE),
    SD_FLE_dva = sd(participant_FLE_dva, na.rm = TRUE),
    
    mean_FLE_ms = mean(participant_FLE_ms, na.rm = TRUE),
    SD_FLE_ms = sd(participant_FLE_ms, na.rm = TRUE),
    
    .groups = "drop"
  ) |>
  arrange(FLE_condition, speed_dva)

FLE_direction_speed_population_stats
```


```{r}
# One value per participant × direction × speed
FLE_speed_plot_data <- FLE_block_speed_data |>
  filter(
    FLE_condition %in% c("petal", "fugal"),
    speed_dva %in% c(10, 15, 20)
  ) |>
  group_by(ID, FLE_condition, speed_dva) |>
  summarise(
    mean_FLE = mean(FLE_dva, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    FLE_condition = factor(
      FLE_condition,
      levels = c("petal", "fugal"),
      labels = c("Petal", "Fugal")
    )
  )

FLE_speed_summary <- FLE_speed_plot_data |>
  group_by(FLE_condition, speed_dva) |>
  summarise(
    n = sum(!is.na(mean_FLE)),
    mean_FLE = mean(mean_FLE, na.rm = TRUE),
    sd_FLE = sd(mean_FLE, na.rm = TRUE),
    se_FLE = sd_FLE / sqrt(n),
    CI_lower = mean_FLE - qt(0.975, df = n - 1) * se_FLE,
    CI_upper = mean_FLE + qt(0.975, df = n - 1) * se_FLE,
    .groups = "drop"
  )

ggplot() +
  # Individual participant trajectories
  geom_line(
    data = FLE_speed_plot_data,
    aes(
      x = speed_dva,
      y = mean_FLE,
      group = interaction(ID, FLE_condition),
      colour = FLE_condition
    ),
    alpha = 0.10,
    linewidth = 0.4
  ) +
  
  # Zero FLE
  geom_hline(
    yintercept = 0,
    colour = "red",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  
  # Confidence intervals
  geom_errorbar(
    data = FLE_speed_summary,
    aes(
      x = speed_dva,
      ymin = CI_lower,
      ymax = CI_upper,
      colour = FLE_condition
    ),
    width = 0.35,
    linewidth = 0.8,
    position = position_dodge(width = 0.5)
  ) +
  
  # Group means
  geom_line(
    data = FLE_speed_summary,
    aes(
      x = speed_dva,
      y = mean_FLE,
      group = FLE_condition,
      colour = FLE_condition
    ),
    linewidth = 1.3
  ) +
  
  geom_point(
    data = FLE_speed_summary,
    aes(
      x = speed_dva,
      y = mean_FLE,
      colour = FLE_condition,
      fill = FLE_condition
    ),
    shape = 21,
    size = 3.5,
    stroke = 0.8
  ) +
  
  scale_x_continuous(
    breaks = c(10, 15, 20)
  ) +
  
  scale_colour_manual(
    name = "Motion direction",
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  
  scale_fill_manual(
    name = "Motion direction",
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  
  labs(
    title = "FLE as a function of speed and motion direction",
    subtitle = "Thick lines show group means and 95% confidence intervals",
    x = "Speed (dva/s)",
    y = "Mean FLE (dva)"
  ) +
  
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

```

```{r}
FLE_low_high_speed <- FLE_speed_plot_data |>
  filter(speed_dva %in% c(10, 20)) |>
  select(ID, FLE_condition, speed_dva, mean_FLE) |>
  pivot_wider(
    names_from = speed_dva,
    values_from = mean_FLE,
    names_prefix = "speed_"
  )

ggplot(
  FLE_low_high_speed,
  aes(
    x = speed_10,
    y = speed_20,
    colour = FLE_condition
  )
) +
  # Zero-reference lines
  geom_hline(
    yintercept = 0,
    colour = "grey40",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  geom_vline(
    xintercept = 0,
    colour = "grey40",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  
  # Participants
  geom_point(
    size = 2.5,
    alpha = 0.75
  ) +
  
  # Separate regression line for each direction
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    linewidth = 1
  ) +
  
  scale_colour_manual(
    name = "Motion direction",
    values = c(
      "Petal" = "#D95F8D",
      "Fugal" = "#377EB8"
    )
  ) +
  
  coord_equal() +
  
  labs(
    title = "Relationship between FLE at the lowest and highest speeds",
    subtitle = "Each point represents one participant",
    x = "Mean FLE at 10 dva/s (dva)",
    y = "Mean FLE at 20 dva/s (dva)"
  ) +
  
  theme_classic(base_size = 13)

# work on this plot
```


:::

# Exploratory Analyses

::: panel-tabset

## Speed x motion direction

```{r}
FLE_speed_model_data <- FLE_block_speed_data |>
  filter(
    FLE_condition %in% c("petal", "fugal"),
    speed_dva %in% c(10, 15, 20)
  ) |>
  mutate(
    ID = factor(ID),
    block = factor(block),
    
    FLE_condition = factor(
      FLE_condition,
      levels = c("fugal", "petal")
    ),
    
    # Centred speed: 0 represents 15 dva/s
    speed_c = speed_dva - 15
  )

FLE_speed_model <- lmer(
  FLE_dva ~ speed_c * FLE_condition +
    (1 + speed_c + FLE_condition | ID) +
    (1 | ID:block),
  data = FLE_speed_model_data,
  REML = FALSE
)

summary(FLE_speed_model)
anova(FLE_speed_model, type = 3)
```

```{r}
speed_slopes <- emtrends(
  FLE_speed_model,
  ~ FLE_condition,
  var = "speed_c"
)

speed_slopes

estimated_FLE <- emmeans(
  FLE_speed_model,
  ~ FLE_condition | speed_c,
  at = list(speed_c = c(-5, 0, 5))
)

estimated_FLE
```

## SD(spatial) = speed x SD(temporal)

```{r}
FLE_SD_by_speed <- FLE_participant_speed_means |>
  group_by(speed_dva) |>
  summarise(
    SD_spatial = sd(participant_FLE_dva, na.rm = TRUE),
    SD_temporal = sd(participant_FLE_ms, na.rm = TRUE),
    predicted_SD_spatial =
      speed_dva * SD_temporal / 1000,
    difference =
      SD_spatial - predicted_SD_spatial,
    .groups = "drop"
  )

FLE_SD_by_speed
```

```{r}
FLE_SD_by_speed |>
  ggplot(
    aes(
      x = speed_dva,
      y = SD_spatial
    )
  ) +
  geom_point(size = 3) +
  geom_line() +
  geom_smooth(
    method = "lm",
    formula = y ~ 0 + x,
    se = FALSE,
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = c(10, 15, 20)) +
  labs(
    x = "Speed (dva/s)",
    y = "Between-participant SD of FLE (dva)",
    title = "Does spatial variability scale with speed?"
  ) +
  theme_classic()
```

```{r}
FLE_SD_by_speed |>
  ggplot(
    aes(
      x = speed_dva,
      y = SD_temporal
    )
  ) +
  geom_point(size = 3) +
  geom_line() +
  scale_x_continuous(breaks = c(10, 15, 20)) +
  labs(
    x = "Speed (dva/s)",
    y = "Between-participant SD of FLE (ms)",
    title = "Temporal variability across speed"
  ) +
  theme_classic()
```

:::
